A Classical Density-Functional Theory for Describing Water Interfaces

### 2011-07-14 paper/project.tex

\subsection{Association free energy}
To the purely repulsive hard-sphere free energy, we add two attractive
energy terms.  The first is the free energy of \emph{association},
which describes the effects of hydrogen bonds.  These are modeled as
four attractive patches (``association sites'') on the surface of the
hard sphere.  These four sites represent the two hydrogens and two
electron lone pairs.  There is an attractive energy
$\epsilon_\textit{assoc}$ when two molecules are arranged such that
the hydrogen of one interacts with the lone pair of the other.  The
volume of this interaction is $\kappa_\textit{assoc}$.  Thus the
association term in the free energy has two free parameters that must
be fit based on experimental data.

The association free energy has the form
\begin{align}
  F_\text{assoc}[n] &= 4 k_BT \int n_0(\xx)
  \left(\ln X(\xx) - \frac{X(\xx)}{2} + \frac12\right) d\xx
\end{align}
where the value of $4$ comes from the four association sites per
molecule, and the functional $X$ is the fraction of assotiation sites
\emph{not} hydrogen-bonded.  The fraction $X$ is determined by the
quadratic equation
\begin{align}
  X(\xx) &= \frac{\sqrt{1 + 8n_0(\xx)\zeta(\xx)\Delta(\xx)} - 1}
  {4 n_0(\xx)\zeta(\xx)\Delta(\xx)}
\end{align}
where $\zeta$ comes from
reference\cite{yu2002fmt-dft-inhomogeneous-associating,
  fu2005vapor-liquid-dft} and describes the inhomogeneity of the fluid
density:
\begin{align}
  \zeta(\xx) &= 1 - \frac{\mathbf{n_{2V}}\cdot\mathbf{n_{2V}}}{n_2^2}
\end{align}
The function $\Delta$ is given... below?

The following contact density for the hard-sphere fluid is in the code
called \texttt{gHS}, and comes
from\cite{yu2002fmt-dft-inhomogeneous-associating,
  fu2005vapor-liquid-dft}, but I don't use it.
\begin{align}
  \Delta &= 4\pi \kappa_\textit{assoc} g^\textit{HS}(\sigma)e^{-\beta
    \epsilon_\textit{assoc}} \\
  g^\textit{HS}(\sigma) &= \frac1{1-n_3}\left(1+\frac14\frac{\zeta n_2}{1-n_3}
  \left(1 + \frac{n_2}{14 (1-n_3)}\right)\right)
\end{align}
\textcolor{red}{Here is paper where I got the association functional:}

Instead, I use
\begin{align}
  \Delta(\xx) &= \kappa_\textit{assoc}
  e^{-\beta\epsilon_\textit{assoc}} g^\textit{SW}_\sigma(\xx) \\
  g^\textit{SW}_\sigma(\xx) &= g^\textit{HS}_\sigma(\xx) +
  \frac{1}{4}\beta\left(\frac{\partial a_1(\xx)}{\partial n_3(\xx)} -
  \frac{\lambda}{3 n3(\xx)}\frac{\partial a_1(\xx)}{\partial \lambda}\right)
\end{align}
% This is equation 77 of gil-villegas-1997-SAFT-VR
which I got from reference\cite{gil-villegas-1997-SAFT-VR}.

>>>>

\subsection{Association free energy}
The first attractive energy term is the association term, which
accounts for hydrogen bonding.  Hydrogen bonds are modeled as four
attractive patches (``association sites'') on the surface of the hard
sphere.  These four sites represent two protons and two electron lone
pairs.  There is an attractive energy $\epsilon_\textit{assoc}$ when
two molecules are arranged such that the proton of one overlaps
with the lone pair of the other.  The volume over which this
interaction occurs is $\kappa_\textit{assoc}$, giving the association
term in the free energy has two empirical parameters that are fit to
experimental data.

The association functional we use is a modified version of that of Yu
and Wu\cite{yu2002fmt-dft-inhomogeneous-associating}.\footnote{We
  should note that Fu and Wu\cite{fu2005vapor-liquid-dft} use almost
  the same functional, but their paper contains errors in the
  association term and is not useful as a reference.}  The functional
of Fu and We includes the effects of density inhomogeneities on the
\emph{contact density} of hard spheres, which conventionally appears
in the form of the density multiplied by the contact value of the
correlation function $g^{HS}_\sigma$, but is based on the SAFT-HS
equation of state\cite{FIXME}, rather than the SAFT-VR equation of
state\cite{gil-villegas-1997-SAFT-VR} which is used in the optimal
SAFT parametrization for water found by Clark \emph{et
  al}\cite{clark2006developing}.

The association free energy for our four-site model has the form
\begin{align}
  F_\text{assoc}[n] &= 4 k_BT \int n_0(\xx)
  \left(\ln X(\xx) - \frac{X(\xx)}{2} + \frac12\right) d\xx
\end{align}
where the value of $4$ comes from the four association sites per
molecule, and the functional $X$ is the fraction of assotiation sites
\emph{not} hydrogen-bonded.  The fraction $X$ is determined by the
quadratic equation
\begin{align}
  X(\xx) &= \frac{\sqrt{1 + 8n_0(\xx)\zeta(\xx)\Delta(\xx)} - 1}
  {4 n_0(\xx)\zeta(\xx)\Delta(\xx)}
\end{align}
where the factor $\zeta$ from
Yu and We\cite{yu2002fmt-dft-inhomogeneous-associating,
  fu2005vapor-liquid-dft} describes the inhomogeneity of the fluid
density, and reduces to unity in the homogeneous limit.
\begin{align}
  \zeta(\xx) &= 1 - \frac{\mathbf{n_{2V}}\cdot\mathbf{n_{2V}}}{n_2^2}
\end{align}
The functional $\Delta$ is a measure of the hydrogen-bonding
probability.  In the SAFT-VR model\cite{gil-villegas-1997-SAFT-VR}, $\Delta$ is given by
\begin{align}
  \Delta &= \kappa_\textit{assoc} g^\textit{SW}_\sigma
  \left(e^{-\beta\epsilon_\textit{assoc}} - 1\right) \\
  g^\textit{SW}_\sigma &= g^\textit{HS}_\sigma +
  \frac{1}{4}\beta\left(\frac{\partial a_1}{\partial n_3} -
  \frac{\lambda}{3 n3}\frac{\partial a_1}{\partial \lambda}\right)
\end{align}
% This is equation 77 of gil-villegas-1997-SAFT-VR
where $g^\textit{SW}_\sigma$ is the correlation function evaluated at
contact for a hard-sphere fluid with a square-well attractive
potential, and $a_1$ and $a_2$ are two terms in the dispersive term of
the free energy.  The correlation function $g^\textit{SW}_\sigma$ is
written as a perturbative correction to the hard-sphere correlation
function, for which we use the functional of Yu and
Wu.\cite{yu2002fmt-dft-inhomogeneous-associating}
\begin{align}
  g_\sigma^{HS} &= \frac{1}{1-n_3}
  +\frac32\frac{\zeta}{(1-n_3)^2}
  + \frac12\frac{n_3^2\zeta}{(1-n_3)^3}
\end{align}
